Dynamic Load Balancing for\\Heterogeneous Systems

### 2011-12-05 paper/paper.tex
Spellchecking on paper.

\begin{abstract}
Leveraging maximimum performance from specialized hardware is one of the
major problems facing computer scientists today.  General-purpose computing
on graphics processing units (GPGPU) is a technique that is becoming
increasingly popular for handling computations typically performed by the
CPU.  However, many GPGPU algorithms place the entire workload on the Graphics
processing unit (GPU) leaving the CPU idle for the duration of the computation.
This paper investigates the effect various load-balancing schemes across the CPU
and GPU have on overall application performance.  Additionally, the discrete and
fused GPU architectures, and their impact on performance, will be investigated.
\end{abstract}

@@

GPU+CPU co-processing hardly a new idea.  Jimenez et al.~\cite{Jimenez2009} present a
dynamic library which can be used in a system to dynamically schedule tasks for GPU
computation across various processes. The method described only looks at optimizing
usage of the GPU across the whole system, rather than optimizing a single application’s
performance. The StarPU system, on the other hand, presents a method of co-processing
and load balancing on a per application basis~\cite{Augonnet2009}. They show that by
using an appropriate model for the load balancing, they are able to achieve improvements
in speed for applications which call for code acceleration multiple times.  We intend
to extend this work to improve the speed of each kernel execution by splitting the
work of that kernel over the capable resources, similar to the schemes used by
OpenMP~\cite{Dagum1998}. Both of these papers address data transfer times as a major cost
of GPU computing. Using the new AMD Fusion architecture, which fuses the CPU and GPU onto
a single die, we may be able to eliminate or greatly reduce this cost, which will impact
the overhead of GPU+CPU co-processing.

@@

Groups of proccessing units, Streaming Multiprocessors (SMs), share a single instruction
dispatch unit and operate in lock step.  The threads themselves run on each SM
in a group, called a warp, typically of 32 threads.  Because of these architectual
features divergent branching in a warp (caused by if-else clauses) greatly degrade application
performance.

The memory architecture for the GPU is very different than that found on the CPU. 
Instead of a large hierarchy of caches to hide memory latency for single thread
performance the latecy to memory is instead hidden through aggressive context
switching.  Newer GPU architectures may contain primative caching schemes.
Each SM on the GPU also contains a local memory memory space which can accessed 
at much lower latency .  This memory space is typically used as a user-managed
cache. The memory hierarchy for the GPU architecture is given by Figure~\ref{fig:gpu_arch}.

@@

For this paper, three different benchmarks were used to assess the
effectiveness of the load balancer.  The first is a reduction benchmark,
in which the sum of variably-sized array is computed.  The reduction benchmark
is unique in that different code is being run on the CPU and the GPU.  The
reason for this is that the GPU cannot efficiently compute a reduction the way
a CPU typically would -- due to the relative innefficiency of backward branches
on the GPU.  A reduction on the GPU is a multi-stage process, in which the
output of one step is used as the input of the next.  Each steam processor
computes the reduction of two elements during each iteration, reducing the amount
of work during each step by a factor of two.  As a result, each step utilizes half
the stream processors of the previous step.  This continues until the reduction
operation is only done by a single stream processor.  A natural consequence of
this algorithm is that not all of the stream processors will be utilized during
the computation.

@@

\section{Load Balancing Schemes}
A naive implementation of CPU+GPU co-processing will in many cases lead to an imbalance
of work for either the CPU or GPU.  To deal with this problem we will investigate the
use of 2 different load balancing schemes in addtion to the performance of a CPU-only or
a GPU-only solution.

@@

To address the issue of a dynamic work load between workgroups for kernels, we also used
a dynamic load-balancing scheme.  In this scheme the worksgroups were grouped into larger
chunks.  Each device is initially allocated one chunk of work to work on.  As each chunk
is retired by the device a new chunk is allocated.  This load-balancing scheme will still
not guarentee optimal performance.  Within a work chunk, a single long-running workgroup
can hold up the entire device, not allowing other kernels to execute on open SMs.

@@

First, the reduction benchmark was run on the discrete GPU platform with a variety of
different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
due to the GPU not utilizing all of its stream processors during a reduction operation, as
discussed previously.  Unfortunately, none of the static load balancing schemes were able
to outperform just the CPU; all the static times were somewhere inbetween the CPU-only and
GPU-only times.  Additionally, the dynamic load balancing scheme performed significantly
worse than all the other schemes, and more than 100\% worse than just the CPU.
100\% worse than just the CPU.

@@

Now, consider Figure \ref{fig:vector_discrete}, where the vector add benchmark was run on
the discrete GPU platform.  In this case, we see that the GPU has nearly a 4x speedup over
just the CPU.  However, similar to the reduction benchmark, the static load balancing schemes
never outperform just the GPU alone, with all of them lying somewhere between the CPU-only
and GPU-only runs.  Also similar to the reduction benchmark, the dynamic load balanching scheme
performs significantly worse than all the other schemes.

@@

Based on the times shown in the previous figures, it is clear that the
static load balancing is ineffective at improving the overal efficiency
of any of the benchmarks tested.  There were a few scenarios in which
a specific static execution ratio achieved meager speedup (<5\%) over
just the GPU, but these situations were scarse, and did not achieve the
speedup we desired.  Additionally, this small amount of speedup would
require a significant amount of tine-tuning by a programmer.

@@

The above results have lead us to believe that the CPU and GPU kernels are
not running truely in parallel as we initially believed.  As such, we used
an AMD profiler to report how long each of the kernels took using the static
load balancer with 95\% of the data on the CPU.  Figure \ref{fig:sequential_discrepancy}
shows the discrepancy between the actual total runtime of the vector add
benchmark and the expected runtime (that is, the sum of the CPU kernel runtime
and the GPU kernel runtime).  If the kernels were truely running in parallel,
we would expect to see a negative discrepancy between the actual runtime and
expected runtime.

@@

\section{Related Work}
Ryoo et al. discuss optimization strategies for GPU versus CPU runtime
systems~\cite{Ryoo2007}.  However, their research is based on a GPU-only
or CPU-only implementation instead of a heterogeneous approach.  The work
presented in ~\cite{Jablin2011} describes a runtime and compiled framework
for improving memory access latency and reducing data transfer overhead
for GPU kernels.  Our work differs from this, because the kernel execution
was still being performed soley on the GPU, whereas our work investigates
kernel execution on multiple devices.

Computation via heterogeneous systems is a relatively new research area for
high performance computing.  The various frameworks and large number of
different accelerator devices can make this quite difficult for application
programmers to manage.  The StarPU system~\cite{Augonnet2009} presents a 
framework to allow for programmers to write numerical codelets and have them
automatically run on a heterogeneous system.  Scheduling work for this system
is one of the most difficult parts of the system, because of the various
overheads for work-sharing schemes. We simplify this work, by looking at a
two-device system, and focus on the performance characteristics of different
scheduling scheme.  Verner et al. use this approach to create a high-throughput
low-lantecy heterogeneous encryption system~\cite{Verner2011}.

>>>>

\begin{abstract}
Leveraging maximum performance from specialized hardware is one of the
major problems facing computer scientists today.  General-purpose computing
on graphics processing units (GPGPU) is a technique that is becoming
increasingly popular for handling computations typically performed by the
CPU.  However, many GPGPU algorithms place the entire workload on the Graphics
processing unit (GPU) leaving the CPU idle for the duration of the computation.
This paper investigates the effect various load-balancing schemes across the CPU
and GPU have on overall application performance.  Additionally, the discrete and
fused GPU architectures, and their impact on performance, will be investigated.
\end{abstract}

@@

GPU+CPU co-processing hardly a new idea.  Jimenez et al.~\cite{Jimenez2009} present a
dynamic library which can be used in a system to dynamically schedule tasks for GPU
computation across various processes. The method described only looks at optimizing
usage of the GPU across the whole system, rather than optimizing a single applications
performance. The StarPU system, on the other hand, presents a method of co-processing
and load balancing on a per application basis~\cite{Augonnet2009}. They show that by
using an appropriate model for the load balancing, they are able to achieve improvements
in speed for applications which call for code acceleration multiple times.  We intend
to extend this work to improve the speed of each kernel execution by splitting the
work of that kernel over the capable resources, similar to the schemes used by
OpenMP~\cite{Dagum1998}. Both of these papers address data transfer times as a major cost
of GPU computing. Using the new AMD Fusion architecture, which fuses the CPU and GPU onto
a single die, we may be able to eliminate or greatly reduce this cost, which will impact
the overhead of GPU+CPU co-processing.

@@

Groups of processing units, Streaming Multiprocessors (SMs), share a single instruction
dispatch unit and operate in lock step.  The threads themselves run on each SM
in a group, called a warp, typically of 32 threads.  Because of these architectural
features divergent branching in a warp (caused by if-else clauses) greatly degrade application
performance.

The memory architecture for the GPU is very different than that found on the CPU. 
Instead of a large hierarchy of caches to hide memory latency for single thread
performance the latency to memory is instead hidden through aggressive context
switching.  Newer GPU architectures may contain privative caching schemes.
Each SM on the GPU also contains a local memory memory space which can accessed 
at much lower latency .  This memory space is typically used as a user-managed
cache. The memory hierarchy for the GPU architecture is given by Figure~\ref{fig:gpu_arch}.

@@

For this paper, three different benchmarks were used to assess the
effectiveness of the load balancer.  The first is a reduction benchmark,
in which the sum of variably-sized array is computed.  The reduction benchmark
is unique in that different code is being run on the CPU and the GPU.  The
reason for this is that the GPU cannot efficiently compute a reduction the way
a CPU typically would -- due to the relative inefficiency of backward branches
on the GPU.  A reduction on the GPU is a multi-stage process, in which the
output of one step is used as the input of the next.  Each steam processor
computes the reduction of two elements during each iteration, reducing the amount
of work during each step by a factor of two.  As a result, each step utilizes half
the stream processors of the previous step.  This continues until the reduction
operation is only done by a single stream processor.  A natural consequence of
this algorithm is that not all of the stream processors will be utilized during
the computation.

@@

\section{Load Balancing Schemes}
A naive implementation of CPU+GPU co-processing will in many cases lead to an imbalance
of work for either the CPU or GPU.  To deal with this problem we will investigate the
use of 2 different load balancing schemes in addition to the performance of a CPU-only or
a GPU-only solution.

@@

To address the issue of a dynamic work load between workgroups for kernels, we also used
a dynamic load-balancing scheme.  In this scheme the workgroups were grouped into larger
chunks.  Each device is initially allocated one chunk of work to work on.  As each chunk
is retired by the device a new chunk is allocated.  This load-balancing scheme will still
not guarantee optimal performance.  Within a work chunk, a single long-running workgroup
can hold up the entire device, not allowing other kernels to execute on open SMs.

@@

First, the reduction benchmark was run on the discrete GPU platform with a variety of
different data sizes and load balancing policies, as shown in Figure \ref{fig:reduce_discrete}.
Interestingly, the CPU outperformed the GPU on this particular benchmark.  This is likely
due to the GPU not utilizing all of its stream processors during a reduction operation, as
discussed previously.  Unfortunately, none of the static load balancing schemes were able
to outperform just the CPU; all the static times were somewhere in between the CPU-only and
GPU-only times.  Additionally, the dynamic load balancing scheme performed significantly
worse than all the other schemes, and more than 100\% worse than just the CPU.
100\% worse than just the CPU.

@@

Now, consider Figure \ref{fig:vector_discrete}, where the vector add benchmark was run on
the discrete GPU platform.  In this case, we see that the GPU has nearly a 4x speedup over
just the CPU.  However, similar to the reduction benchmark, the static load balancing schemes
never outperform just the GPU alone, with all of them lying somewhere between the CPU-only
and GPU-only runs.  Also similar to the reduction benchmark, the dynamic load balancing scheme
performs significantly worse than all the other schemes.

@@

Based on the times shown in the previous figures, it is clear that the
static load balancing is ineffective at improving the overall efficiency
of any of the benchmarks tested.  There were a few scenarios in which
a specific static execution ratio achieved meager speedup (<5\%) over
just the GPU, but these situations were scarce, and did not achieve the
speedup we desired.  Additionally, this small amount of speedup would
require a significant amount of tine-tuning by a programmer.

@@

The above results have lead us to believe that the CPU and GPU kernels are
not running truly in parallel as we initially believed.  As such, we used
an AMD profiler to report how long each of the kernels took using the static
load balancer with 95\% of the data on the CPU.  Figure \ref{fig:sequential_discrepancy}
shows the discrepancy between the actual total runtime of the vector add
benchmark and the expected runtime (that is, the sum of the CPU kernel runtime
and the GPU kernel runtime).  If the kernels were truly running in parallel,
we would expect to see a negative discrepancy between the actual runtime and
expected runtime.

@@

\section{Related Work}
Ryoo et al. discuss optimization strategies for GPU versus CPU runtime
systems~\cite{Ryoo2007}.  However, their research is based on a GPU-only
or CPU-only implementation instead of a heterogeneous approach.  The work
presented in ~\cite{Jablin2011} describes a runtime and compiled framework
for improving memory access latency and reducing data transfer overhead
for GPU kernels.  Our work differs from this, because the kernel execution
was still being performed solely on the GPU, whereas our work investigates
kernel execution on multiple devices.

Computation via heterogeneous systems is a relatively new research area for
high performance computing.  The various frameworks and large number of
different accelerator devices can make this quite difficult for application
programmers to manage.  The StarPU system~\cite{Augonnet2009} presents a 
framework to allow for programmers to write numerical codelets and have them
automatically run on a heterogeneous system.  Scheduling work for this system
is one of the most difficult parts of the system, because of the various
overheads for work-sharing schemes. We simplify this work, by looking at a
two-device system, and focus on the performance characteristics of different
scheduling scheme.  Verner et al. use this approach to create a high-throughput
low-latency heterogeneous encryption system~\cite{Verner2011}.
